---
title: Testing Clustering methods.
author: Hayley Wragg
date: "21st Nov 2023"
output:
  html_document:
    keep_md: yes
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

# Install package

Load the "ndimclusteringR" package

```{r}
devtools::install("../ndimclusteringR/R")
library("ndimclusteringR")
```

# Source test

Load the functions for testing

```{r}
source("test_clust_kmeans.R")
```

# Step by step

If all the steps are included then they are performed in the following order.

1\.
Crop data to complete cases and lowest se.

2\.
Convert data to angles.

3\.
Perform PCA on angle data.

4\.
Cluster on the PCA

-   Set cluster centres
-   Converge cluster centres
-   if min then rerun for different number of clusters and choose set with min-aic.

### Test 1

Run the tests with:

-   No PCA
-   basic k-means clustering. k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 2.

```{r}
pc_type <- "No_pca"
np <- 1
d <- 30
clust_typ <- "basic"
space_typ <- "regular"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 2

Run the tests with:

-   No PCA
-   basic k-means clustering. k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 4.

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "basic"
space_typ <- "regular"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 3

Run the tests with:

-   No PCA
-   basic k-means clustering. k is set to the number of pathways.
-   angles calculated before pca and before clustering.
-   number of pathways 2.

```{r}
pc_type <- "No_pca"
np <- 1
d <- 30
clust_typ <- "basic"
space_typ <- "angle"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 4

Run the tests with:

-   No PCA
-   basic k-means clustering. k is set to the number of pathways.
-   angles calculated before clustering.
-   number of pathways 4.

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "basic"
space_typ <- "angle"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 5

Run the tests with:

-   No PCA
-   min-aic k-means clustering. Maximum k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 2.

```{r}
pc_type <- "No_pca"
np <- 1
d <- 30
clust_typ <- "min"
space_typ <- "regular"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 6

Run the tests with:

-   No PCA
-   min-aic k-means clustering. Maximum k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 4.

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "min"
space_typ <- "regular"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 7

Run the tests with:

-   No PCA
-   min k-means clustering. Maximum k is set to the number of pathways.
-   angles calculated before clustering.
-   number of pathways 2.

```{r}
pc_type <- "No_pca"
np <- 1
d <- 30
clust_typ <- "min"
space_typ <- "angle"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

### Test 8

Run the tests with:

-   No PCA
-   min-aic k-means clustering. Maximum k is set to the number of pathways.
-   angles calculated before clustering.
-   number of pathways 4.

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "min"
space_typ <- "angle"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ)
```

## Consider the location of the centroids.

The previous tests used centroids assigned randomly within the ranges on each axis.
Instead we will initialise the centroids using randomly selected points from out dataset.

### Test 9

Run the tests with:

-   No PCA
-   basic k-means clustering. k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "basic"
space_typ <- "regular"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 10

Run the tests with: - No PCA - basic k-means clustering.
k is set to the number of pathways.
- angles calculated before clustering.
- number of pathways 4.
- Centroids assigned using points

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "basic"
space_typ <- "angle"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 11

Run the tests with:

-   PCA with prcomp
-   basic k-means clustering. k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "prcomp"
np <- 3
d <- 30
clust_typ <- "basic"
space_typ <- "regular"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 12

Run the tests with:

-   PCA with prcomp
-   basic k-means clustering. k is set to the number of pathways.
-   angles calculated before PCA and clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "prcomp"
np <- 3
d <- 30
clust_typ <- "basic"
space_typ <- "angle"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 13

Run the tests with:

-   No PCA
-   min-aic k-means clustering. k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "min"
space_typ <- "regular"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 14

Run the tests with:

-   No PCA
-   min-aic k-means clustering. k is set to the number of pathways.
-   angles calculated before clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "No_pca"
np <- 3
d <- 30
clust_typ <- "min"
space_typ <- "angle"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 15

Run the tests with:

-   PCA with prcomp
-   min-aic k-means clustering. k is set to the number of pathways.
-   no angles calculated before clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "prcomp"
np <- 3
d <- 30
clust_typ <- "min"
space_typ <- "regular"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```

### Test 16

Run the tests with:

-   PCA with prcomp
-   min-aic k-means clustering. k is set to the number of pathways.
-   angles calculated before PCA and clustering.
-   number of pathways 4.
-   Centroids assigned using points

```{r}
pc_type <- "prcomp"
np <- 3
d <- 30
clust_typ <- "min"
space_typ <- "angle"
how_cents <- "points"
test_clust_kmeans_function(d,
                           num_path = np,
                           pc_type = pc_type,
                           clust_typ = clust_typ,
                           space_typ = space_typ,
                           how_cents = how_cents)
```
